# Introduction {-}

## Getting Started {#getting-started .unnumbered}

Before we continue, make sure you have all the software you need for this book:

-   **R**: If you don't have R installed already, you may be reading the wrong book; I assume a basic familiarity with R throughout this book.
    If you'd like to learn how to use R, I'd recommend my [*R for Data Science*](https://r4ds.had.co.nz/) which is designed to get you up and running with R with a minimum of fuss.

-   **RStudio**: RStudio is a free and open source integrated development environment (IDE) for R.
    While you can write and use Shiny apps with any R environment (including R GUI and [ESS](http://ess.r-project.org)), RStudio has some nice features specifically for authoring, debugging, and deploying Shiny apps.
    We recommend giving it a try, but it's not required to be successful with Shiny or with this book.
    You can download RStudio Desktop from <https://www.rstudio.com/products/rstudio/download>

## System Requirements {#system-requirements .unnumbered}

This book was written in [RStudio](http://www.rstudio.com/ide/) using [bookdown](http://bookdown.org/).

The [website](http://jeffamaxey.github.io/soa-exam-pa) is hosted with [Github Pages](https://github.com/), and automatically updated after every commit by [Github Actions](https://github.com/features/actions).
The complete source is available from [GitHub](https://github.com/jeffamaxey/soa-exa-pa).

This version of the book was built with `r R.version.string` and the following packages:

```{r, echo = FALSE, results="asis", messages=FALSE, warning=FALSE}
library(tidyverse)
library(tibble)
library(kableExtra)

pkgs <- sessioninfo::package_info(pkgs = "attached", dependencies = FALSE)
df <- tibble(
  package = pkgs$package,
  version = pkgs$ondiskversion,
  source = gsub("@", "\\\\@", pkgs$source)
)
df %>%
  kbl(df, format = "html", caption = "R Packages") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = F,
    position = "center",
    fixed_thead = T
  )
```


```{r, echo = FALSE}
ruler <- function(width = getOption("width")) {
  x <- seq_len(width)
  y <- dplyr::case_when(
    x %% 10 == 0 ~ as.character((x %/% 10) %% 10),
    x %% 5 == 0  ~ "+",
    TRUE         ~ "-"
  )
  cat(y, "\n", sep = "")
  cat(x %% 10, "\n", sep = "")
}
```

```{r, include = FALSE}
ruler()
```

